Methods paper for erosion

### 2017-12-04 docs/methods/erosion_methods.tex
Changes to latex methods paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods\label{s:method}} 
There are three main numerical methods that need to be developed:~the
boundary integral equation solver for
equation~\eqref{eqn:completed_BIE}, computing the quantities of interest
in Section~\ref{sec:qois}, and updating the shape of the eroding bodies
with equation~\eqref{eqn:bodyEvolution}.  So that we can form stable
simulations with long time horizons, we use numerical methods that
achieve spectral accuracy in space, fourth-order accuracy in time, and
maintain an equispaced discretization of the bodies.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial Discretization}
\label{sec:spatial_discretization}
Each body is represented with $N_\iin$ points as the Fourier series
\begin{align*}
  \xx^\ell(\alpha) = \sum_{k = -N_\iin /2}^{N_\iin /2-1} 
    \hat{x}_k^\ell e^{ik\alpha}, \quad \ell=1,\ldots,M.
\end{align*}
The exterior geometry is represented with a Fourier series using
$N_\out$ points.  Derivatives of the shape are computed in Fourier space
\begin{align*}
  \frac{d\xx(\alpha)}{d\alpha} = 
    \sum_{k = -N_\out /2}^{N_\out /2-1} ik\hat{x}_k e^{ik\alpha}.
\end{align*}

@@

In this section, we also describe our time stepping method and two
regularization terms that are necessary to control the curvature of the
bodies and to maintain spectral accuracy in space.

Once the shear stress is computed, it is used to update the shape of the
bodies according to the normal flow~\eqref{eqn:bodyEvolution}.  To
eliminate tangling and stretching of the mesh, we use a {\thL}
formulation~\cite{hou-low-she1994} which is described in
Section~\ref{sec:thetaL}.

We use a quasi-static approximation so that the velocity field
instantaneously changes with the changing geometry.  Therefore, we can
alternate between the two main aspects of our simulation
\begin{enumerate}
  \item Given a geometry, compute the shear stress of along the boundary
    of each of the bodies.
  \item Given the shear stress, update the geometry according to the
    erosion model.
\end{enumerate}

\vsp{2}
To handle boundary evolution, we use the {\thL}    method, which offers
certain advantages in the ability to stabilize fluid-structure
interaction problems \cite{hou-low-she1994}. Throughout this section, we
use the following conventions: the bodies are parametrized in the CCW
direction and the normal vector points into the bodies (our out of the
fluid).

In order to regularize boundary evolution, we modify the normal velocity
in Eq.~\eqref{eqn:bodyEvolution} as follows
\begin{equation}
\label{eqn:Vn}
\Vn = \atau + \eps \mean{\atau} \left( \frac{L \kappa}{2\pi} - 1 \right)
%\Vn = \atau + \frac{\eps}{\log \left(2\pi/L \right)} \left(\kappa - \frac{2 \pi}{L} \right)
\end{equation}
Here, $\kappa$ is the curvature, $\eps$ is a small parameter, and $L$ is the total arc length of the body. The second term on the right acts to smooth bodies that have high-frequency oscillations. Since we have subtracted the mean curvature, $2\pi/L$, this term is area-preserving, so that the only material loss is due to the shear stress. 
\todo[inline]{Modify the text here.}
We chose the prefactor $\log^{-1} \left(2\pi/L \right)$, so that, for 2D Stokes flow, this term scales the same as shear stress as $L \to 0$. Thus, the relative strength of the smoothing is always the same.

@@

For simplicity, we introduce the variables
\begin{align}
\label{eq:M}
& M = - 2 \pi \mean{\thalpha \Vn} \\
\label{eq:N}
& N = \frac{2 \pi}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right) \\
\label{eq:elfun}
& \elfun = \frac{2 \pi  \mean{\atau}}{L}
%\frac{4\pi^2}{L^{2} \log \left(2\pi/L \right)}
\end{align} 
Then Eq.~\eqref{eqn:thetaevo0} can be rewritten as
\begin{equation}
\label{eqn:thetaevo1}
\pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + N
\end{equation}
\todo[inline]{Change $N$ to something else $\mathcal{N}$}
Note that the $\elfun$ depends only on $L(t)$, and so the highest order
derivative in Eq.~\eqref{eqn:thetaevo1} is {\em linear} in the variable $\alpha$. This is the key to the {\thL} method, and allows us to use very stable, implicit methods for the stiff term. We will evolve $\theta$ in spectral space, and so introduce the transforms
\begin{align}
& \theta(\alpha,t) = \sum_{k = -\infty}^{\infty} \thhat_k e^{ i k \alpha} \\
& N(\alpha,t)  = \sum_{k = -\infty}^{\infty} \hat{N}_k e^{ i k \alpha}
\end{align}
Equation~\eqref{eqn:thetaevo1} gives a set of ODEs for the coefficients
\begin{equation}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \hat{N}_k
\end{equation}
We introduce the integrating factor
\begin{equation}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right)
\end{equation}
Then the evolution equations for $L$ and $\theta$ can be recast as the coupled system of ODEs
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = M \\
\label{eqn:thetaevo2}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \hat{N}_k
\end{align}
We solve system \eqref{eqn:Levo2}--\eqref{eqn:thetaevo2} with a combination of an explicit, second-order Runge-Kutta method (RK2) and trapezoid rule to compute the integrating factor, $\mu_k$. There is not a unique second-order Runge-Kutta method, and we choose the {\em midpoint} version to maximize the smoothing effect of the curvature term.

At the first stage of RK2 (the half step), we discretize Eqs.~\eqref{eqn:Levo2}--\eqref{eqn:thetaevo2} as
\begin{align}
\label{eq:L12}
& \frac{L^{(n+1/2)} - L^{(n)}}{\Dt/2} = M^{(n)} \\
& \frac{ \mu_k^{(n+1/2)} \thhat_k^{(n+1/2)} - \thhat_k^{(n)}}{\Dt/2} = \hat{N}_k^{(n)}
\end{align}
where the superscript indicates the time step. For simplicity, we have chosen the constant of integration in the exponent of Eq.~\eqref{eq:mu} so that $\mu_k^{(n)} = 1$. At the next stage of RK2, we must compute $M^{(n+1/2)}$ and $\hat{N}_k^{(n+1/2)}$, which requires a call to the Stokes solver. Hence, we must compute compute $\mu_k^{(n+1/2})$, so that $\thhat_k^{(n+1/2)}$ can be recovered and the geometry at the mid-step can be computed. To compute $\mu_k^{(n+1/2)}$, we use the left-end-point rule (first order, consistent with first forward Euler step of RK2)
\begin{equation}
\mu_k^{(n+1/2)} = \exp \left( \epsilon k^2 \frac{\Dt}{2} \elfun^{(n)}\right)
%\exp \left( \frac{1}{4} \eps k^2 \Dt \left( \elfun^{(n)} + \elfun^{(n+1/2)} \right) \right)
\end{equation}
\todo[inline]{modify text below}
Notice in Eq.~\eqref{eq:elfun}, that $\elfun$ depends only on $L(t)$. Hence, we just need to update $L^{(n+1/2)}$ first via Eq.~\eqref{eq:L12}, and we have what we need to compute $\mu_k^{(n+1/2)}$. It is instructive to recast as
\begin{equation}
\thhat_k^{(n+1/2)} = \left( \thhat_k^{(n)} + \frac{1}{2} \Dt \, \hat{N}_k^{(n)} \right)
\exp \left( -\epsilon k^2 \frac{\Dt}{2} \elfun^{(n)}\right)
%\exp \left( - \frac{1}{4} \eps k^2 \Dt \left( \elfun^{(n)} + \elfun^{(n+1/2)} \right) \right)
\end{equation}
so that the smoothing effect of the exponential is obvious. With $\thhat_k^{(n+1/2)}$ computed, it is now possible to compute $M^{(n+1/2)}$ and $\hat{N}_k^{(n+1/2)}$.

Next, at the second stage of RK2, we discretize as
\begin{align}
& \frac{L^{(n+1)} - L^{(n)}}{\Dt} = M^{(n+1/2)} \\
& \frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \thhat_k^{(n)}}{\Dt} = \mu_k^{(n+1/2)} \hat{N}_k^{(n+1/2)}
\end{align}
We again compute the integrating factor, $\mu_k^{(n+1)}$ via the ( insert integration rule), giving
\begin{align}
\thhat_k^{(n+1)} = 
& \thhat_k^{(n)}
\exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) +  \\
& \Dt \, \hat{N}_k^{(n+1/2)}
\exp \left( - \frac{1}{4} \eps k^2 \Dt \left( 3 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right)
%& \thhat_k^{(n)}
%\exp \left( - \frac{1}{2} \eps k^2 \Dt \left( \elfun^{(n)} + \elfun^{(n+1)} \right) \right) +  \\
%& \Dt \, \hat{N}_k^{(n+1/2)}
%\exp \left( - \frac{1}{4} \eps k^2 \Dt \left( \elfun^{(n+1/2)} + \elfun^{(n+1)} \right) \right)
\end{align}
CHECK

\vsp{5}
\todo[inline]{Artificial Diffusion} 
\todo[inline]{Gaussian Filter}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods\label{s:method}} 
There are three main numerical methods that need to be developed:~the
boundary integral equation solver for
equation~\eqref{eqn:completed_BIE}, computing the quantities of interest
in Section~\ref{sec:qois}, and updating the shape of the eroding bodies
with equation~\eqref{eqn:bodyEvolution}.  So that we can form stable
simulations with long time horizons, we use numerical methods that
achieve spectral accuracy in space, second-order accuracy in time, and
maintain an equispaced discretization of the bodies.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial Discretization}
\label{sec:spatial_discretization}
Each body is represented with $N_\iin$ points as the Fourier series
\begin{align*}
  \xx^\ell(\alpha) = \FourierSum \hat{x}_k^\ell e^{ik\alpha}, \quad \ell=1,\ldots,M.
\end{align*}
The exterior geometry is represented with a Fourier series using
$N_\out$ points.  Derivatives of the shape are computed in Fourier space
\begin{align*}
  \frac{d\xx(\alpha)}{d\alpha} = \FourierSum ik\hat{x}_k e^{ik\alpha}.
\end{align*}

@@

% BRYAN'S ORIGINAL TEXT
%In this section, we also describe our time stepping method and two regularization terms that are necessary to control the curvature of the bodies and to maintain spectral accuracy in space.

%Once the shear stress is computed, it is used to update the shape of the bodies according to the normal flow~\eqref{eqn:bodyEvolution}.  To eliminate tangling and stretching of the mesh, we use a {\thL}formulation~\cite{hou-low-she1994} which is described in Section~\ref{sec:thetaL}.

%We use a quasi-static approximation so that the velocity field instantaneously changes with the changing geometry.  Therefore, we can alternate between the two main aspects of our simulation
%\begin{enumerate}
%\item Given a geometry, compute the shear stress of along the boundary of each of the bodies.
%\item Given the shear stress, update the geometry according to the erosion model.
%\end{enumerate}

To handle boundary evolution, we use the {\thL}    method, which offers
certain advantages in the ability to stabilize fluid-structure
interaction problems \cite{hou-low-she1994}. Throughout this section, we
use the following conventions: the bodies are parametrized in the CCW
direction and the normal vector points into the bodies (our out of the
fluid).

In order to regularize boundary evolution, we modify the normal velocity
in Eq.~\eqref{eqn:bodyEvolution} as follows
\begin{equation}
\label{eqn:Vn}
\Vn = \atau + \eps \mean{\atau} \left( \frac{L \kappa}{2\pi} - 1 \right)
%\Vn = \atau + \frac{\eps}{\log \left(2\pi/L \right)} \left(\kappa - \frac{2 \pi}{L} \right)
\end{equation}
Here, $\kappa$ is the curvature, $\eps$ is a small parameter, and $L$ is the total arc length of the body, and $\mean{\cdot}$ indicates a surface mean. The second term on the right acts to smooth bodies with high-frequency oscillations by introducing a {\em curvature}-dependent component in the interface velocity. Since a closed body has a mean curvature of $\mean{\kappa} = 2\pi / L$, the term in parentheses has mean zero. Thus, this term is area-preserving, so that the only material loss is due to the shear stress itself. In addition, we have scaled this term on the $\mean{\atau}$, the average absolute stress, so that it is always proportional to the true source of erosion $\atau$, even as bodies vanish (reword).

@@

For simplicity, we introduce the variables
\begin{align}
\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vn} \\
\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right) \\
\label{eq:elfun}
& \elfun = \frac{2 \pi  \mean{\atau}}{L}
%\frac{4\pi^2}{L^{2} \log \left(2\pi/L \right)}
\end{align} 
Then Eq.~\eqref{eqn:thetaevo0} can be rewritten as
\begin{equation}
\label{eqn:thetaevo1}
\pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm
\end{equation}
Note that the highest-order, i.e.~stiff, term in Eq.~\eqref{eqn:thetaevo1} is {\em quasi-linear}. This is the key to the {\thL} method, and allows us to use very stable, implicit methods for this stiff term. We will evolve $\theta$ in spectral space, and so introduce the transforms
\begin{align}
& \theta(\alpha,t) = \FourierSum \thhat_k e^{ i k \alpha} \\
& \NLterm(\alpha,t)  = \FourierSum \hat{\NLterm}_k e^{ i k \alpha}
\end{align}
Equation~\eqref{eqn:thetaevo1} gives a set of ODEs for the coefficients
\begin{equation}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \hat{\NLterm}_k
\end{equation}
We introduce the integrating factor
\begin{equation}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right)
\end{equation}
Then the evolution equations for $L$ and $\theta$ can be recast as the coupled system of ODEs
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm \\
\label{eqn:thetaevo2}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \hat{\NLterm}_k
\end{align}
We solve system \eqref{eqn:Levo2}--\eqref{eqn:thetaevo2} with an explicit, second-order Runge-Kutta method (RK2). In particular, we choose the {\em midpoint} RK2 method so as to maximize the smoothing effect of the curvature term.

At the first stage of RK2 (the half step), we discretize Eqs.~\eqref{eqn:Levo2}--\eqref{eqn:thetaevo2} as
\begin{align}
\label{eq:L12}
& \frac{L^{(n+1/2)} - L^{(n)}}{\Dt/2} = \Mterm^{(n)} \\
& \frac{ \mu_k^{(n+1/2)} \thhat_k^{(n+1/2)} - \thhat_k^{(n)}}{\Dt/2} = \hat{\NLterm}_k^{(n)}
\end{align}
where the superscript indicates the time step. For simplicity, we have chosen the constant of integration in the exponent of Eq.~\eqref{eq:mu} so that $\mu_k^{(n)} = 1$. At the next stage of RK2, we must compute $\Mterm^{(n+1/2)}$ and $\hat{\NLterm}_k^{(n+1/2)}$, which requires a call to the Stokes solver. Hence, we must compute $\mu_k^{(n+1/2})$, so that $\thhat_k^{(n+1/2)}$ can be recovered and the geometry at the mid-step can be computed. To compute $\mu_k^{(n+1/2)}$, we use the left-end-point rule (first order, consistent with first forward Euler step of RK2)
\begin{equation}
\mu_k^{(n+1/2)} = \exp \left( \epsilon k^2 \frac{\Dt}{2} \elfun^{(n)}\right)
\end{equation}
It is instructive to recast as
\begin{equation}
\thhat_k^{(n+1/2)} = \left( \thhat_k^{(n)} + \frac{1}{2} \Dt \, \hat{\NLterm}_k^{(n)} \right)
\exp \left( -\epsilon k^2 \frac{\Dt}{2} \elfun^{(n)}\right)
\end{equation}
so that the smoothing effect of the exponential is obvious. With $\thhat_k^{(n+1/2)}$ computed, it is now possible to compute $\Mterm^{(n+1/2)}$ and $\hat{\NLterm}_k^{(n+1/2)}$.

Next, at the second stage of RK2, we discretize as
\begin{align}
& \frac{L^{(n+1)} - L^{(n)}}{\Dt} = \Mterm^{(n+1/2)} \\
& \frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \thhat_k^{(n)}}{\Dt} = \mu_k^{(n+1/2)} \hat{\NLterm}_k^{(n+1/2)}
\end{align}
We again compute the integrating factor, $\mu_k^{(n+1)}$ via the ( insert integration rule), giving
\begin{align}
\thhat_k^{(n+1)} = 
& \thhat_k^{(n)}
\exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) +  \\
& \Dt \, \hat{\NLterm}_k^{(n+1/2)}
\exp \left( - \frac{1}{4} \eps k^2 \Dt \left( 3 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right)
\end{align}

\newpage
\todo[inline]{Artificial Diffusion} 
\todo[inline]{Gaussian Filter}
